---
title: "Hello World!"
author: "Claudia Tanaka"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set( message=FALSE, warning=FALSE )
options(scipen=999) # "Desliga" notação científica. Para "ligar" use scipen=0


library(tidyverse) # para trabalhar com tabelas
library(arrow)
library(gt)

theme_set(theme_light())
theme_update(
  panel.grid.minor = element_blank(),
  plot.title = element_text(size = 12, colour = "gray30", face = "bold"),
  plot.subtitle = element_text(face = 'italic', colour = "gray50", size = 10),
  plot.caption = element_text(colour = "gray50", hjust=0, size = 8),
  legend.title = element_blank(),
)
```


<br>

## Auxiliary Tables

```{r}
# Tabela auxiliar de códigos e classificações de países
ta.countries <- wbstats::wb_countries() |> 
  select(iso3c, country, region, income_level)
```


```{r}
ta.who <- readxl::read_excel("_datasets/GHED_data.XLSX", sheet=2,.name_repair="universal") |> 
  select(code=variable.code, name=variable.name, unit, currency) |> 
  separate(code, into=c("code","extras"), sep="_") |> 
  filter(is.na(extras))
```



```{r}
ta.who.raw <- readxl::read_excel("_datasets/GHED_data.XLSX", sheet=1) |> 
  filter(year != 2022)
```


<br>

## Health expenditure

<br>

Data downloaded from [WHO Global Health Expenditure Update 2023](https://apps.who.int/nha/database)

Last updated: December 3rd, 2023

Note: Year 2021 added. Data for 2022 are preliminary and subject to revision. 



```{r}
who.tt <- ta.who.raw |> 
  select(
    iso3c=code, income, year, 
    che_gdp,        # Current Health Expenditure (CHE) as % Gross Domestic Product (GDP)
    pop,            # Population (in thousands)
    # Amounts in constant (2021) US$ per capita:
    che_usd2021,    # Current Health Expenditure (CHE)
    gdp_usd2021_pc, # Gross Domestic Product (GDP)
    gge_usd2021_pc  # General government expenditure (GGE), 
  ) |> 
  mutate(gdp_usd2021 = pop*gdp_usd2021_pc,
         gge_usd2021 = pop*gge_usd2021_pc) |> 
  inner_join(ta.countries, join_by(iso3c))
```

<br>

Inspect

```{r}
colSums(is.na(who.tt)) |> as.data.frame() |> rownames_to_column() |> rename(coluna=1,nulos=2) |> 
  filter(nulos>0) |> arrange(desc(nulos))
```

<br>

```{r}
who.tt |> filter(is.na(che_usd2021)) |> 
  summarise(n=n(), .by=c(iso3c, country, region)) |> 
  arrange(desc(n))
```

<br>


```{r}
who.tt |> filter(is.na(pop)) |> 
  summarise(n=n(), .by=c(iso3c, country, region)) |> 
  arrange(desc(n))
```


<br>

```{r}
who.tt <- who.tt |> na.omit()
```


<br>

```{r}
range(who.tt$year) 
```


<br>

```{r}
who.tt |> filter(year > max(year)-5) |> 
  summarise(
    across(c(gdp_usd2021, gge_usd2021, che_usd2021), ~round(sum(.x, na.rm=T)/1e+6,1)), 
    pop = round(sum(pop, na.rm=T),0),
    n.countries = n(),
    .by=year
  ) |> 
  arrange(year)
```


<br>

**Sources**   


```{r}
who.sources <- ta.who.raw |> 
  select(
    iso3c=code, income, year, 
    gghed_usd2021,  # Domestic General Government (GGHE-D)
    pvtd_usd2021,   # Domestic Private (PVT-D)
    ext_usd2021,    # External (EXT)
  ) |> 
  inner_join(ta.countries, join_by(iso3c)) |> 
  pivot_longer(gghed_usd2021:ext_usd2021, names_to="source", values_drop_na=T) |> 
  filter(value > 0) |> 
  mutate(source = str_split_i(source, "_", 1)) |> 
  left_join(ta.who |> select(code, name), join_by(source==code))
```


<br>

Inspect

```{r}
colSums(is.na(who.sources)) |> as.data.frame() |> rownames_to_column() |> rename(coluna=1,nulos=2) |> 
  filter(nulos>0) |> arrange(desc(nulos))
```

<br>

Summarize


```{r}
who.sources |> filter(year==max(year)) |> 
  summarise(n.countries = n(),
            value = round(sum(value, na.rm=T),0), 
            .by=c(year, name)) |> 
  mutate(tt = sum(value))
```

<br>


```{r}
who.sources |> filter(year > max(year)-5) |> 
  summarise(n.countries = n_distinct(iso3c),
            value = round(sum(value, na.rm=T)), 
            .by=c(year)) |> 
  arrange(year)
```



<br>

## Financing schemes:

Current health expenditure by financing schemes


```{r}
who.schemes <- ta.who.raw |> 
  select(
    iso3c=code, income, year, 
    hf1_usd2021,  # Government schemes and compulsory contributory health care financing schemes
    hf2_usd2021,  # Voluntary health care payment schemes
    hf3_usd2021,  # Household out-of-pocket payments (OOPS)
    hf4_usd2021,  # Rest of the world financing schemes (non-resident)
    hfnec_usd2021 # Unspecified financing schemes (n.e.c.)
  ) |> 
  inner_join(ta.countries, join_by(iso3c)) |>  
  pivot_longer(hf1_usd2021:hfnec_usd2021, names_to="scheme", values_drop_na=T) |>
  filter(value > 0) |>
  mutate(scheme = str_split_i(scheme, "_", 1)) |> 
  left_join(ta.who |> select(code, name), join_by(scheme==code))
```


<br>

Inspect

```{r}
colSums(is.na(who.schemes)) |> as.data.frame() |> rownames_to_column() |> rename(coluna=1,nulos=2) |> 
  filter(nulos>0) |> arrange(desc(nulos))
```

<br>

Summarize

```{r}
who.schemes |> filter(year==max(year)) |> 
  summarise(n.countries = n(),
            value = round(sum(value, na.rm=T),0), 
            .by=c(year, scheme)) |> 
  mutate(tt = sum(value))
```

<br>


```{r}
who.schemes |> filter(year > max(year)-5) |> 
  summarise(n.countries = n_distinct(iso3c),
            tt = round(sum(value, na.rm=T)), 
            .by=c(year)) |> 
  arrange(year)
```

<br>

**Government schemes and compulsory contributory health care:**
hf11_usd2021	  Government schemes
hf12_usd2021	  Compulsory contributory health insurance schemes
hf13_usd2021	  Compulsory Medical Saving Accounts (CMSA)
hf1nec_usd2021	Unspecified government schemes and compulsory contributory schemes (n.e.c.)


**Compulsory contributory health insurance schemes**
hf121_usd2021	  Social health insurance schemes
hf122_usd2021	  Compulsory private insurance schemes
hf12nec_usd2021	Unspecified compulsory contributory health insurance schemes (n.e.c.)


**Voluntary health care payment schemes**
hf21_usd2021	  Voluntary health insurance schemes
hf22_usd2021	  NPISH financing schemes (including development agencies)
hf23_usd2021	  Enterprise financing schemes
hf2nec_usd2021	Unspecified voluntary health care payment schemes (n.e.c.)





<br>


Data downloaded from [OECD Data Explorer](https://data-explorer.oecd.org/)

Read .csv and write to .parquet 

```{r}
# Health expenditure and financing
## DATAFLOW  OECD.ELS.HD:DSD_SHA@DF_SHA(1.0)
## df.health_exp <- read_csv("_datasets/OECD.ELS.HD,DSD_SHA@DF_SHA,1.0+all.csv",name_repair="universal")
#
# df.health_exp |>
#   group_by(TIME_PERIOD) |>
#   arrow::write_dataset("_datasets/OECD_HEALTH_EXP/")
```


<br>

Read from .parquet


```{r}
oecd.hlth.exp <- open_dataset("_datasets/OECD_HEALTH_EXP/") |>
  select(-c(STRUCTURE, STRUCTURE_ID, STRUCTURE_NAME, ACTION,
            Measure, MEASURE, Currency, Decimals, DECIMALS,
            FINANCING_SCHEME_REV, Revenues.of.financing.schemes,
            FACTOR_PROVISION, Factor.of.provision, Observation.value, Base.period,
            Asset.type, ASSET_TYPE, PRICE_BASE,
            Time.period,
            OBS_STATUS, OBS_STATUS2, OBS_STATUS3, Observation.status.3,
            UNIT_MULT,
            FREQ, Frequency.of.observation)) |>
  rename(iso3c=REF_AREA, country=Reference.area, year=TIME_PERIOD,
         health_function=7) |>
  mutate(year = as.integer(year)) |>
  rename_with(tolower)
```


<br>

### Inspect

```{r}
glimpse(oecd.hlth.exp)
```


<br>

```{r}
colSums(is.na(collect(oecd.hlth.exp))) |> as.data.frame() |> rownames_to_column() |> 
  rename(coluna=1,nulos=2) |> 
  filter(nulos>0) |> arrange(desc(nulos))
```



```{r}
unique(collect(oecd.hlth.exp)$price.base)
```



```{r}
oecd.hlth.exp |> 
  summarise(n=n(), .by=c(unit_measure, unit.of.measure)) |> 
  arrange(unit_measure) |> 
  collect()
```



```{r}
oecd.hlth.exp |> 
  summarise(n=n(), .by=c(financing_scheme, financing.scheme)) |> 
  arrange(financing_scheme) |> 
  collect()
```




```{r}
oecd.hlth.exp |> 
  summarise(n=n(), .by=c(health_function, health.function)) |> 
  arrange(health_function) |> 
  collect()
```




```{r}
oecd.hlth.exp |> 
  summarise(n=n(), .by=c(mode_provision, mode.of.provision)) |> 
  arrange(mode_provision) |> 
  collect()
```


```{r}
oecd.hlth.exp |> 
  summarise(n=n(), .by=c(provider, health.care.provider)) |> 
  arrange(provider) |> 
  collect()
```



<br>

### Summarize


```{r plt_exp1, echo=FALSE, fig.width=6, fig.asp=.6}
oecd.hlth.exp |> 
  filter(financing_scheme=="_T", mode_provision=="_T", health_function=="_T", provider=="_T",
         unit_measure=="USD_PPP", price.base=="Constant prices") |>
  select(-c(unit_measure, unit.of.measure, price.base)) |> 
  select(year, country, obs_value) |> collect() |> 
  summarise(value = sum(obs_value), .by=year) |> 
  ggplot(aes(x=year, y=value, group=1)) +
  geom_line(color="blue") + geom_point(color="blue") +
  scale_y_continuous(labels=scales::label_number(scale=1/1000), limits=c(0,NA)) +
  labs(
    x=NULL, y=NULL,
    title="Total Health Expenditure, USD billions",
    subtitle="PPP converted at constant prices",
    caption="Source: OECD.Stat, National Accounts - Main aggregates. Extracted June 2023, Paris."
  )
```

<br>

```{r plt_exp_count, echo=FALSE, fig.width=6, fig.asp=.6}
oecd.hlth.exp |> 
  filter(financing_scheme=="_T", mode_provision=="_T", health_function=="_T", provider=="_T",
         unit_measure=="USD_PPP", price.base=="Constant prices") |>
  select(-c(unit_measure, unit.of.measure, price.base)) |> 
  select(year, country, obs_value) |> collect() |> 
  summarise(value = n_distinct(country), .by=year) |> 
  ggplot(aes(x=year, y=value, group=1)) +
  geom_line(color="blue") + geom_point(color="blue") +
  scale_y_continuous(labels=scales::label_number()) +
  labs(
    x=NULL, y=NULL,
    title="Number of countries",
    caption="Source: OECD.Stat, National Accounts - Main aggregates. Extracted June 2023, Paris."
  )
```

<br>

**Health Expenditure by Financing Scheme, 2022**    
*USD Millions PPP converted at constant prices*


```{r echo=FALSE}
oecd.hlth.exp |> 
  filter(year == 2022, 
         financing_scheme!="_T", mode_provision=="_T", health_function=="_T", provider=="_T",
         unit_measure=="USD_PPP", price.base=="Constant prices") |> 
  summarise(
    countries = n_distinct(iso3c),
    health_usd = round(sum(obs_value)),
    .by=c(financing.scheme)) |> 
  collect() |> 
  mutate(pct = round(health_usd/sum(health_usd)*100, 1)) |> 
  arrange(desc(health_usd))
```


*Source: OECD.Stat, National Accounts - Main aggregates. Extracted June 2023, Paris.*

<br>

Create a dataset with total health expenditure, percent of GDP and health spend per capita per country and year.

Create a dataset with financing schemes detailed by country and year.

<br>

## Coverage

```{r}
# DATAFLOW  OECD.ELS.HD:DSD_HEALTH_PROT@DF_HEALTH_PROT(1.0)
# Social protection
# HIC - Health insurance coverage
oecd.cvrg <- read_csv("_datasets/OECD.HI_coverage.csv", name_repair="universal") |>
  select(-c(STRUCTURE, STRUCTURE_ID, STRUCTURE_NAME, ACTION,
            FREQ, Frequency.of.observation,
            Measure, MEASURE, UNIT_MEASURE, Unit.of.measure,
            INSURANCE_TYPE,
            Time.period, Observation.value,
            OBS_STATUS,
            UNIT_MULT, Unit.multiplier,
            Decimals, DECIMALS)) |>
  rename(iso3c=REF_AREA, country=Reference.area, year=TIME_PERIOD) |>
  mutate(year = as.integer(year)) |>
  rename_with(tolower)
```


<br>

### Inspect

```{r}
glimpse(oecd.cvrg)
```



```{r}
range(oecd.cvrg$year)
```



```{r}
oecd.cvrg |> summarise(n=n(), .by=c(insurance.type))
```

<br>

### Summarize

**Mean health coverage, 2022**      
*as percent of population*

```{r echo=FALSE}
oecd.cvrg |> 
  filter(year == 2022) |> 
  summarise(
    countries = n_distinct(iso3c),
    pct = round(mean(obs_value),1),
    .by=c(insurance.type)) |> 
  arrange(desc(pct))
```
*Source: OECD, OECD Health Statistics 2023. July 2023.*

<br>

Must add population data to get weighted average of coverage by insurance type.