---
title: "Hello World!"
author: "Claudia Tanaka"
date: "`r Sys.Date()`"
output: github_document
---

```{r setup, include = FALSE, cache = FALSE}
knitr::opts_chunk$set( message=FALSE, warning=FALSE )
options(scipen=999) # "Desliga" notação científica. Para "ligar" use scipen=0


library(tidyverse) # para trabalhar com tabelas
library(arrow)
library(gt)

theme_set(theme_light())
theme_update(
  panel.grid.minor = element_blank(),
  plot.title = element_text(size = 12, colour = "gray30", face = "bold"),
  plot.subtitle = element_text(face = 'italic', colour = "gray50", size = 10),
  plot.caption = element_text(colour = "gray50", hjust=0, size = 8),
  legend.title = element_blank(),
)
```


<br>

## Health expenditure

Data downloaded from [OECD Data Explorer](https://data-explorer.oecd.org/)

Read .csv and write to .parquet 

```{r}
# Health expenditure and financing
## DATAFLOW  OECD.ELS.HD:DSD_SHA@DF_SHA(1.0)
## df.health_exp <- read_csv("_datasets/OECD.ELS.HD,DSD_SHA@DF_SHA,1.0+all.csv",name_repair="universal")
#
# df.health_exp |>
#   group_by(TIME_PERIOD) |>
#   arrow::write_dataset("_datasets/OECD_HEALTH_EXP/")
```


<br>

Read from .parquet


```{r}
oecd.hlth.exp <- open_dataset("_datasets/OECD_HEALTH_EXP/") |>
  select(-c(STRUCTURE, STRUCTURE_ID, STRUCTURE_NAME, ACTION,
            Measure, MEASURE, Currency, Decimals, DECIMALS,
            FINANCING_SCHEME_REV, Revenues.of.financing.schemes,
            FACTOR_PROVISION, Factor.of.provision, Observation.value, Base.period,
            Asset.type, ASSET_TYPE, PRICE_BASE,
            Time.period,
            OBS_STATUS, OBS_STATUS2, OBS_STATUS3, Observation.status.3,
            UNIT_MULT,
            FREQ, Frequency.of.observation)) |>
  rename(iso3c=REF_AREA, country=Reference.area, year=TIME_PERIOD,
         health_function=7) |>
  mutate(year = as.integer(year)) |>
  rename_with(tolower)
```


<br>

### Inspect

```{r}
glimpse(oecd.hlth.exp)
```


<br>

```{r}
colSums(is.na(collect(oecd.hlth.exp))) |> as.data.frame() |> rownames_to_column() |> 
  rename(coluna=1,nulos=2) |> 
  filter(nulos>0) |> arrange(desc(nulos))
```



```{r}
unique(collect(oecd.hlth.exp)$price.base)
```



```{r}
oecd.hlth.exp |> 
  summarise(n=n(), .by=c(unit_measure, unit.of.measure)) |> 
  arrange(unit_measure) |> 
  collect()
```



```{r}
oecd.hlth.exp |> 
  summarise(n=n(), .by=c(financing_scheme, financing.scheme)) |> 
  arrange(financing_scheme) |> 
  collect()
```




```{r}
oecd.hlth.exp |> 
  summarise(n=n(), .by=c(health_function, health.function)) |> 
  arrange(health_function) |> 
  collect()
```




```{r}
oecd.hlth.exp |> 
  summarise(n=n(), .by=c(mode_provision, mode.of.provision)) |> 
  arrange(mode_provision) |> 
  collect()
```


```{r}
oecd.hlth.exp |> 
  summarise(n=n(), .by=c(provider, health.care.provider)) |> 
  arrange(provider) |> 
  collect()
```



<br>

### Summarize


```{r plt_exp1, echo=FALSE, fig.width=6, fig.asp=.6}
oecd.hlth.exp |> 
  filter(financing_scheme=="_T", mode_provision=="_T", health_function=="_T", provider=="_T",
         unit_measure=="USD_PPP", price.base=="Constant prices") |>
  select(-c(unit_measure, unit.of.measure, price.base)) |> 
  select(year, country, obs_value) |> collect() |> 
  summarise(value = sum(obs_value), .by=year) |> 
  ggplot(aes(x=year, y=value, group=1)) +
  geom_line(color="blue") + geom_point(color="blue") +
  scale_y_continuous(labels=scales::label_number(scale=1/1000), limits=c(0,NA)) +
  labs(
    x=NULL, y=NULL,
    title="Total Health Expenditure, USD billions",
    subtitle="PPP converted at constant prices",
    caption="Source: OECD.Stat, National Accounts - Main aggregates. Extracted June 2023, Paris."
  )
```

<br>

```{r plt_exp_count, echo=FALSE, fig.width=6, fig.asp=.6}
oecd.hlth.exp |> 
  filter(financing_scheme=="_T", mode_provision=="_T", health_function=="_T", provider=="_T",
         unit_measure=="USD_PPP", price.base=="Constant prices") |>
  select(-c(unit_measure, unit.of.measure, price.base)) |> 
  select(year, country, obs_value) |> collect() |> 
  summarise(value = n_distinct(country), .by=year) |> 
  ggplot(aes(x=year, y=value, group=1)) +
  geom_line(color="blue") + geom_point(color="blue") +
  scale_y_continuous(labels=scales::label_number()) +
  labs(
    x=NULL, y=NULL,
    title="Number of countries",
    caption="Source: OECD.Stat, National Accounts - Main aggregates. Extracted June 2023, Paris."
  )
```

<br>

**Health Expenditure by Financing Scheme, 2022**    
*USD Millions PPP converted at constant prices*


```{r echo=FALSE}
oecd.hlth.exp |> 
  filter(year == 2022, 
         financing_scheme!="_T", mode_provision=="_T", health_function=="_T", provider=="_T",
         unit_measure=="USD_PPP", price.base=="Constant prices") |> 
  summarise(
    countries = n_distinct(iso3c),
    health_usd = sum(obs_value),
    .by=c(financing.scheme)) |> 
  collect() |> 
  mutate(pct = round(health_usd/sum(health_usd)*100, 1)) |> 
  arrange(desc(health_usd))
```


*Source: OECD.Stat, National Accounts - Main aggregates. Extracted June 2023, Paris.*

<br>

Create a dataset with total health expenditure, percent of GDP and health spend per capita per country and year.

Create a dataset with financing schemes detailed by country and year.

<br>

## Coverage

```{r}
# DATAFLOW  OECD.ELS.HD:DSD_HEALTH_PROT@DF_HEALTH_PROT(1.0)
# Social protection
# HIC - Health insurance coverage
oecd.cvrg <- read_csv("_datasets/OECD.HI_coverage.csv", name_repair="universal") |>
  select(-c(STRUCTURE, STRUCTURE_ID, STRUCTURE_NAME, ACTION,
            FREQ, Frequency.of.observation,
            Measure, MEASURE, UNIT_MEASURE, Unit.of.measure,
            INSURANCE_TYPE,
            Time.period, Observation.value,
            OBS_STATUS,
            UNIT_MULT, Unit.multiplier,
            Decimals, DECIMALS)) |>
  rename(iso3c=REF_AREA, country=Reference.area, year=TIME_PERIOD) |>
  mutate(year = as.integer(year)) |>
  rename_with(tolower)
```


<br>

### Inspect

```{r}
glimpse(oecd.cvrg)
```



```{r}
range(oecd.cvrg$year)
```



```{r}
oecd.cvrg |> summarise(n=n(), .by=c(insurance.type))
```

<br>

### Summarize

**Mean health coverage, 2022**      
*as percent of population*

```{r echo=FALSE}
oecd.cvrg |> 
  filter(year == 2022) |> 
  summarise(
    countries = n_distinct(iso3c),
    pct = mean(obs_value),
    .by=c(insurance.type)) |> 
  arrange(desc(pct))
```
*Source: OECD, OECD Health Statistics 2023. July 2023.*

<br>

Must add population data to get weighted average of coverage by insurance type.